Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2022-09-28 paper/paper-infer.tex
integrate Clara's comments

\title{Inferring disease architecture and predictive ability\\with LDpred2-auto}
\author{Florian Priv\'e,$^{\text{1,}*}$ Bogdan Pasaniuc,$^{\text{2,3,4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6,7}}$}

@@

Most traits and diseases are heritable, if not all. 
What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
Knowing how much a phenotype is heritable and how much it is polygenic can teach a lot about it.
Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SuSiE (per-variant $p$), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,wang2020simple,zeng2021widespread,johnson2021estimation}.

These two parameters are useful e.g.\ to determine how well we can predict a phenotype from using genetic variants alone \cite[]{daetwyler2008accuracy,dudbridge2013power}. 
Such genetic predictors are called polygenic scores (PGS), and are getting closer to be included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
LDpred2 is a very competitive polygenic score method that can directly builds PGS using summary statistics results from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
LDpred2 uses the (SNP) heritability and polygenicity as parameters of its model. In LDpred2-auto, it can directly estimate these parameters from the data, making it applicable even when no validation data is available for tuning these two model hyper-parameters \cite[]{prive2020ldpred2}.

Here we extend LDpred2-auto to make it a highly reliable method for estimating (local) $h^2$, (per-variant) $p$, and $\alpha$.
So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without needing an independent test set.
An overview of what LDpred2-auto can provide is presented in Figure \ref{fig:overview}.
Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
We call this new preferred set of 1,444,196 SNPs HapMap3+.

@@

For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}; we use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP-heritability $h^2$, the polygenicity $p$ (i.e. the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
This function first picks a proportion $p$ of causal variants at random, sample effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scale the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix. Finally, some Gaussian noise is added so that the full phenotype has a variance of 1. 
Then, a GWAS is run to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a subset of 20,000) using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
Finally, we run the new LDpred2-auto model (Methods) with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled). 
LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$.

LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha}).
Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$), and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
The $\alpha$ estimate becomes unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
Moreover, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).

We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a thresholding corresponding to the prevalence.
Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$). The only difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2012better}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.

@@

Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}). Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature.

We then investigate using two alternative LD references (Methods).
Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:alt_LD}).

We then investigate using the extended set of HapMap3 variants proposed here (Methods). As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}) instead of 4--12 hours before.

@@

In this section, we use the extended set of variants constructed here, HapMap3+. 
We recall that we have defined 431 independent LD blocks for this set (Methods).
We compute per-block (local) $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by one block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).

Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure \ref{fig:median_local_h2}).
The outlier block explaining a much larger heritability contains the HLA region.
Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a negative correlation of -5.5\% with the probabilities of being causal.
To verify that this does not correspond to regions of low LD, we compute LD scores; the probabilities above have a correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.

@@

Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M European individuals \cite[]{yengo2022saturated}.
Using the method developed in \cite{prive2021using}, we first infer the ancestry proportions of individuals included in the meta-analysis as 81.9\% from N.W. Europe, 9.5\% from E. Europe, 6.5\% from Finland, 1.5\% of Ashkenazi ancestry, 0.3\% from S.W. Europe, and 0.2\% from W. Africa. 
We can therefore use the same N.W. European LD with these GWAS summary statistics. 
Note that we use the HapMap3+ set of 1,444,196 SNPs here, but that we use only 1,013,499 SNPs for the GWAS meta-analysis (those overlapping and passing quality control).

Intercepts from LD Score regression are increasing with sample size: 1.01 (SE: 0.007) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, as expected \cite[]{loh2018mixed}.
SNP heritability estimates are of 65.5\% (SE: 2.8), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively. And of 60.2\% [95\% CI: 57.4, 63.1], 62.9\% [61.7, 64.0], and 54.1\% [53.8, 54.4] with LDpred2-auto, respectively.
The reduced heritability estimation from the meta-analysis can be partly explained by three reasons: a reduced number of overlapping SNPs, more heterogeneity in the individuals from the meta-analysis compared to within the UK Biobank only, and less effect of assortative mating (which is strong in the UK Biobank due to the presence of many spouses, and probably contributes to overestimating the heritability of height there).
As expected, predictive performance $r^2$ are increasing with sample size, with 30.0\% [29.2, 30.8], 42.7\% [42.2, 43.1], and 46.9\% [46.8, 47.1], respectively.
Note that the first two estimates (from UKBB) are slightly overestimated, and that the last one is based on a smaller number of overlapping SNPs.
We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only (46.9\% versus 54.1\%).
Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.1\% [1.0, 1.4], 2.2\% [2.0, 2.4], and 5.7\% [5.4, 6.0], consistent with results of simulations with a large polygenicity (p=10\%). Therefore, we estimate that height has at least 50,000 causal variants.
We also identify 1744 SNPs with a probability of being causal larger than 95\%.
As for $\alpha$ estimates from LDpred2-auto, they remain highly consistent, with -0.72 [-0.76, -0.68], -0.74 [-0.76, -0.71], and -0.78 [-0.81, -0.75], respectively.
Finally, we compute per-annotation heritability estimates to investigate functional enrichment. We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).

@@

LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the (local) SNP-heritability, polygenicity (and per-variant probabilities of being causal), and the selection-related parameter $\alpha$.
It can also infer the predictive performance $r^2$ of the resulting PGS (for the same genetic ancestry as the GWAS used for training).
We remind readers that it can also be used to infer the incertainty of individual polygenic scores \cite[]{ding2022large}.
Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf generally providing much lower predictive performance than LDpred2-grid or LDpred2-auto \cite[]{prive2020ldpred2}.
We also obtain widespread signatures of selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
However, when looking at the heritability enrichment of several functional annotations for height, we obtain much smaller magnitudes than stratified LD Score regression \cite[]{finucane2018heritability}. For example, \cite{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x (Figure \ref{fig:enrichment}).

Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 40\% larger to offer a better coverage of the genome.
This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
Using this new set also improves predictive performance here, particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
It means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.

Our proposed method have limitations. First, when power is low, estimates of $\alpha$ and $p$ become less reliable. However, estimates of $h^2$ and $r^2$ seem always reliable, except for maybe height for which they are probably overestimated.
We think this is due to assortative mating making the estimates too large in the presence of many spouses such as for the UK Biobank [TODO: CITATIONS? YENGO2022? OTHERS?]. 
Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics; future work could focus on correcting this issue. [TODO: OTHERS?]

>>>>

\title{Inferring disease architecture and predictive ability\\with LDpred2-auto}
\author{Florian Priv\'e,$^{\text{1,}*}$ Clara Albi\~nana,$^{\text{1}}$ Bogdan Pasaniuc,$^{\text{2,3,4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6,7}}$}

@@

Most traits and diseases are heritable. 
What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
Understanding how heritable and polygenic a phenotype is can provide valuable clinical insight and inform future research efforts.
Therefore, many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome. 
These methods include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$ used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$ that can inform about negative selection, \cite{schoech2019quantification}), SuSiE (per-variant $p$ used for fine-mapping, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).

As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone. 
Such genetic predictors are called polygenic scores (PGS), and are getting closer to being included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
LDpred2 is a widely used polygenic score method that can directly build PGS using summary statistics results from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
LDpred2 is a Bayesian approach that uses the SNP heritability $h^2$ and polygenicity $p$ as parameters of its model. In LDpred2-auto, it can directly estimate these parameters from the data, making it applicable even when no validation data is available for tuning these two model hyper-parameters \cite[]{prive2020ldpred2}.

Here we extend LDpred2-auto to make it a highly reliable method for estimating $h^2$ (global and local), $p$ (also per-variant for fine-mapping purposes), and $\alpha$.
So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without needing an independent test set.
An overview of what LDpred2-auto can provide is presented in Figure \ref{fig:overview}.
Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
We call this new preferred set of 1,444,196 SNPs HapMap3+.

@@

[TODO: INCLUDE SIMPLIFIED VERSIONS OF FIG S1-S3 HERE?]

For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}. We use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP-heritability $h^2$, the polygenicity $p$ (i.e. the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
This function first picks a proportion $p$ of causal variants at random, sample effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scale the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix. Finally, some Gaussian noise is added so that the full phenotype has a variance of 1. 
Then, a GWAS is run to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a subset of 20,000) using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
Finally, we run the new LDpred2-auto model (Methods) with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled, in which case it prevents effect sizes from changing sign without going through 0 first). 
LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale).
[TODO: EXPLAIN FILTERING OF CHAINS]

First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha}).
Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$), and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
The $\alpha$ estimate can become unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
Then, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
We recall that calibrated per-variant probabilities of being causal can be used for fine-mapping purposes  \cite[]{wang2020simple}.
Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).

We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a threshold corresponding to the prevalence.
Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$). The only difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2012better}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.

@@

Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}). Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').

Then, to investigate whether estimates from LDpred2-auto are robust to misspecifications, we test using two alternative LD references (Methods).
Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:alt_LD}).

We then investigate using the extended set of HapMap3 variants proposed here, which includes \textasciitilde40\% more variants than the set of HapMap3 variants recommended to use for LDpred2 (Methods). As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}). 
This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}) instead of 4--12 hours before.

@@

In this section, we use the extended set of variants constructed here, HapMap3+, for which we define 431 independent LD blocks (Methods).
We compute per-block (local) $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by one block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).

Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure \ref{fig:median_local_h2}).
The outlier block explaining a much larger heritability contains the HLA region.
Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; interestingly, these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a negative correlation of -5.5\% with the probabilities of being causal.
To verify that this does not correspond to regions of low LD, we compute LD scores; the probabilities above have a correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.

@@

Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M European individuals [TODO: ADD NUMBER OF VARIANTS] \cite[]{yengo2022saturated}.
Using the method developed in \cite{prive2021using}, we first infer the ancestry proportions of individuals included in the meta-analysis as 81.9\% from N.W. Europe, 9.5\% from E. Europe, 6.5\% from Finland, 1.5\% of Ashkenazi ancestry, 0.3\% from S.W. Europe, and 0.2\% from W. Africa. 
We can therefore use the same N.W. European LD with these GWAS summary statistics. 
Note that we use the HapMap3+ set of 1,444,196 SNPs here, but that we use only 1,013,499 SNPs for the GWAS meta-analysis (those overlapping and passing quality control).

Intercepts from LD Score regression are increasing with sample size: 1.01 (SE: 0.007) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, as expected \cite[]{loh2018mixed}.
SNP heritability estimates are of 65.5\% (SE: 2.8), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively. And of 60.2\% [95\% CI: 57.4, 63.1], 62.9\% [61.7, 64.0], and 54.1\% [53.8, 54.4] with LDpred2-auto, respectively.
The reduced heritability estimation from the meta-analysis can be partly explained by three reasons: a reduced number of overlapping SNPs, more heterogeneity in the individuals from the meta-analysis compared to within the UK Biobank only, and less effect of assortative mating (which is strong in the UK Biobank due to the presence of many spouses, and probably contributes to overestimating the heritability of height there).
As expected, predictive performance $r^2$ are increasing with sample size, with 30.0\% [29.2, 30.8], 42.7\% [42.2, 43.1], and 46.9\% [46.8, 47.1], respectively.
Note that the first two estimates (from UKBB) are slightly overestimated, and that the last one is based on a smaller number of overlapping SNPs.
We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only (46.9\% versus 54.1\%).
Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.1\% [1.0, 1.4], 2.2\% [2.0, 2.4], and 5.7\% [5.4, 6.0], consistent with results of simulations with a large polygenicity (p=10\%). 
[TODO: RERUN FOR HEIGHT WITH JUMP\_SIGN]
Therefore, we estimate that height has at least 50,000 causal variants.
We also identify 1744 SNPs with a probability of being causal larger than 95\% (fine-mapping).
As for $\alpha$ estimates from LDpred2-auto, they remain highly consistent, with -0.72 [-0.76, -0.68], -0.74 [-0.76, -0.71], and -0.78 [-0.81, -0.75], respectively.
Finally, we compute per-annotation heritability estimates to investigate functional enrichment. We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).

@@

LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP-heritability (global and local), polygenicity (and per-variant probabilities of being causal), and the selection-related parameter $\alpha$.
It can also infer the predictive performance $r^2$ of the resulting PGS (for the same genetic ancestry as the GWAS used for training).
We remind readers that it can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf generally providing much lower predictive performance than LDpred2-grid or LDpred2-auto \cite[]{prive2020ldpred2}.
We also obtain widespread signatures of selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
However, when looking at the heritability enrichment of several functional annotations for height, we obtain much smaller magnitudes than stratified LD Score regression \cite[]{finucane2018heritability}. For example, \cite{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x.

Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 40\% larger to offer a better coverage of the genome.
This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
Using this new set also improves predictive performance here, particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
It means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.

Our proposed method has limitations. First, when power is low, estimates of $\alpha$ and $p$ become less reliable. However, estimates of $h^2$ and $r^2$ seem always reliable, except for maybe height for which they are probably overestimated.
We think this is likely due to assortative mating \cite[]{border2022assortative,yengo2022saturated}. 
Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics; future work could focus on correcting this issue. [TODO: OTHERS?]

Nevertheless, LDpred2-auto users can now get much from running a single method only. The reliable estimates provided by LDpred2-auto are very encouraging to further extend LDpred2-auto in multiple directions. 
As future research directions, we are interested in using LDpred2-auto for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}, as well as extending it to learn from functional annotations \cite[]{zhang2021improved,marquez2021incorporating}.
